---
title: "EBSmoothr Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

load_ebsmoothr <- function() {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("../../EBSmoothr", quiet = TRUE, export_all = FALSE)
    return(invisible(TRUE))
  }
  if (requireNamespace("EBSmoothr", quietly = TRUE)) {
    library(EBSmoothr)
    return(invisible(TRUE))
  }
  stop("Install EBSmoothr or pkgload before rendering this document.")
}

load_ebsmoothr()
```

# Purpose

`EBSmoothr` provides empirical-Bayes smoothers for Gaussian means problems.
The package currently exposes two prior families:

- `LGP`: a one-dimensional local Gaussian-process smoother with a TMB backend.
- `Matern`: a one-dimensional or two-dimensional exact Gaussian Matern smoother
  built from an INLA/SPDE precision representation, with an optional INLA
  PC-prior backend.

In both cases, the user supplies:

- observations `x`,
- known standard errors `s`, and
- either a precomputed setup object or a generator built from locations.

The returned fitter follows the `ebnm` convention:

```{r}
library(EBSmoothr)

fit_fun <- ebnm_LGP_generator(
  LGP_setup(t = seq(0, 1, length.out = 10))
)

str(formals(fit_fun))
```

# Typical Workflow

The package is designed around a two-step workflow.

1. Construct a smoothing family or setup object.
2. Generate an `ebnm`-compatible fitting function.
3. Fit the model to `x` and `s`.
4. Inspect posterior summaries such as `posterior$mean`, `posterior$var`, and `fitted_g`.

For L-GP:

```{r}
t <- seq(0, 8, length.out = 100)
setup <- LGP_setup(t = t, p = 2, num_knots = 25, betaprec = 0)
fit_fun <- ebnm_LGP_generator(setup, link = "identity")
```

For Matern:

```{r}
locations <- seq(0, 8, length.out = 100)
fit_fun <- ebnm_Matern_generator(locations = locations, link = "identity")
```

# Choosing Between L-GP and Matern

Use `LGP` when:

- the domain is one-dimensional,
- a spline-like local basis is natural, and
- you want explicit control over the polynomial trend basis and knot count.

Use `Matern` when:

- the domain is one-dimensional or two-dimensional,
- a spatial Gaussian-process interpretation is preferred, or
- you want an exact Gaussian empirical-Bayes fit based on an INLA/SPDE
  precision representation over a mesh.

If you instead want Matern hyperparameters to be selected under a PC prior,
pass `pc.penalty` to `ebnm_Matern_generator()` and the fitter will switch to
an INLA backend.

# About the Link Function

The package is framed around Gaussian observations.

- `LGP` supports `"identity"` and `"log"`.
- `Matern` currently supports `"identity"` only, because the exact Gaussian
  implementation is built around the closed-form marginal likelihood for the
  normal-means case.

# Minimal Simulated Example

```{r}
set.seed(1)

t <- seq(0, 10, length.out = 120)
truth <- sin(t) + 0.25 * cos(2 * t)
s <- rep(0.25, length(t))
x <- truth + rnorm(length(t), sd = s)

setup <- LGP_setup(t = t, p = 2, num_knots = 30, betaprec = 0)
fit_fun <- ebnm_LGP_generator(setup, link = "identity")
fit <- fit_fun(x, s)

head(fit$posterior)
fit$fitted_g
```

```{r}
plot(t, x, pch = 16, cex = 0.5, col = "grey40",
     main = "L-GP smoothing example",
     xlab = "t", ylab = "observed value")
lines(t, truth, col = "firebrick", lwd = 2, lty = 2)
lines(t, fit$posterior$mean, col = "navy", lwd = 2)
```

# Output Structure

Both L-GP and Matern fitters return an object with class `c("list", "ebnm")`.
Common components include:

- `posterior`: posterior means, variances, and second moments.
- `fitted_g`: the fitted prior-family parameterization.
- `fitted_beta`: the fitted or fixed global coefficient(s).
- `g_init`: the initialization used for the family.
- `log_likelihood`: the main model-comparison quantity returned by the fitter.

The exact additional components depend on the backend. For example, the Matern
fit exposes the mesh and a latent spatial-field summary; `inla_result` is
`NULL` in exact mode and stores the INLA fit object in `pc.penalty` mode.
