---
title: "EBSmoothr Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

load_ebsmoothr <- function() {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("../../EBSmoothr", quiet = TRUE, export_all = FALSE)
    return(invisible(TRUE))
  }
  if (requireNamespace("EBSmoothr", quietly = TRUE)) {
    library(EBSmoothr)
    return(invisible(TRUE))
  }
  stop("Install EBSmoothr or pkgload before rendering this document.")
}

load_ebsmoothr()
```

# Purpose

`EBSmoothr` provides empirical-Bayes smoothers for Gaussian means problems.
The package currently exposes two prior families:

- `LGP`: a one-dimensional local Gaussian-process smoother with a TMB backend.
- `Matern`: a one-dimensional or two-dimensional exact Gaussian Matern smoother
  built from an INLA/SPDE precision representation, with an optional INLA
  PC-prior backend.

At the top level, the package is organized around two collaborator-facing
workflows:

- Known observation standard errors `s`: use `ebnm_LGP_generator()` or
  `ebnm_Matern_generator()` to build an `ebnm`-compatible fitter.
- Unknown observation standard errors: use `eb_smoother()` with `s = NULL`
  to learn one common observation noise SD.

`eb_smoother()` also accepts known `s`, but this overview focuses on the two
main entry points above so collaborators can quickly choose the right API.

# Entry-Point Guide

## Known `s`: generator workflow

When standard errors are known and you want an `ebnm`-compatible fit, the
package follows the generator pattern:

```{r}
library(EBSmoothr)

fit_fun <- ebnm_LGP_generator(
  LGP_setup(t = seq(0, 1, length.out = 10))
)

str(formals(fit_fun))
```

The resulting fitter has the usual `ebnm` signature `function(x, s, ...)` and
returns an object with class `c("list", "ebnm")`.

## Unknown `s`: high-level smoothing workflow

When standard errors are unknown, `eb_smoother()` is the main entry point. It
learns one common observation noise SD and returns an `eb_smoother_fit` object
with compact `print()` / `summary()` methods.

```{r}
set.seed(1)

loc <- seq(0, 8, length.out = 100)
truth <- sin(loc) + 0.25 * cos(2 * loc)
x <- truth + rnorm(length(loc), sd = 0.2)

fit_spatial <- eb_smoother(
  x = x,
  s = NULL,
  family = "matern",
  locations = loc
)

fit_baseline <- eb_smoother(
  x = x,
  s = NULL,
  family = "constant",
  beta_prec = 0
)

fit_spatial
summary(fit_baseline)
```

# Typical Workflows

The package therefore has two common setup patterns.

1. Known `s`: construct a family-specific setup object if needed, generate an
   `ebnm`-compatible fitting function, and fit the model to `x` and `s`.
2. Unknown `s`: call `eb_smoother()` directly with `s = NULL`, choose the
   family, and inspect the resulting `eb_smoother_fit`.

For the known-`s` generator workflow, a typical L-GP setup looks like:

```{r}
t <- seq(0, 8, length.out = 100)
setup <- LGP_setup(t = t, p = 2, num_knots = 25, betaprec = 0)
fit_fun <- ebnm_LGP_generator(setup, link = "identity")
```

For the known-`s` generator workflow, a typical Matern setup looks like:

```{r}
locations <- seq(0, 8, length.out = 100)
fit_fun <- ebnm_Matern_generator(locations = locations, link = "identity")
```

# Choosing Between L-GP and Matern

Use `LGP` when:

- the domain is one-dimensional,
- a spline-like local basis is natural, and
- you want explicit control over the polynomial trend basis and knot count.

Use `Matern` when:

- the domain is one-dimensional or two-dimensional,
- a spatial Gaussian-process interpretation is preferred, or
- you want an exact Gaussian empirical-Bayes fit based on an INLA/SPDE
  precision representation over a mesh.

If you instead want Matern hyperparameters to be selected under a PC prior,
pass `pc.penalty` to `ebnm_Matern_generator()` and the fitter will switch to
an INLA backend.

If standard errors are unknown, the same family choice carries over to
`eb_smoother()` via `family = "lgp"` or `family = "matern"`, while
`family = "constant"` provides the Gaussian baseline with no spatial
structure.

# About the Link Function

The package is framed around Gaussian observations.

- `LGP` supports `"identity"` and `"log"`.
- `Matern` currently supports `"identity"` only, because the exact Gaussian
  implementation is built around the closed-form marginal likelihood for the
  normal-means case.

# Minimal Known-`s` Example

```{r}
set.seed(2)

t <- seq(0, 10, length.out = 120)
truth <- sin(t) + 0.25 * cos(2 * t)
s <- rep(0.25, length(t))
x <- truth + rnorm(length(t), sd = s)

setup <- LGP_setup(t = t, p = 2, num_knots = 30, betaprec = 0)
fit_fun <- ebnm_LGP_generator(setup, link = "identity")
fit <- fit_fun(x, s)

head(fit$posterior)
fit$fitted_g
```

```{r}
plot(t, x, pch = 16, cex = 0.5, col = "grey40",
     main = "L-GP smoothing example",
     xlab = "t", ylab = "observed value")
lines(t, truth, col = "firebrick", lwd = 2, lty = 2)
lines(t, fit$posterior$mean, col = "navy", lwd = 2)
```

# Output Structure

The generator workflows return objects with class `c("list", "ebnm")`.
Common components include:

- `posterior`: posterior means, variances, and second moments.
- `fitted_g`: the fitted prior-family parameterization.
- `fitted_beta`: the fitted or fixed global coefficient(s).
- `g_init`: the initialization used for the family.
- `log_likelihood`: the main model-comparison quantity returned by the fitter.

The exact additional components depend on the backend. For example, the Matern
fit exposes the mesh and a latent spatial-field summary; `inla_result` is
`NULL` in exact mode and stores the INLA fit object in `pc.penalty` mode.

`eb_smoother()` instead returns an object with class
`c("eb_smoother_fit", "list")`. In addition to the stored posterior summaries,
these fits expose:

- `fitted_noise_sd` for the learned-noise workflow,
- `family`, `backend`, and `noise_mode` metadata,
- `fitted_g` and `fitted_beta`, and
- compact `print()` / `summary()` methods for collaborator-facing inspection.
