---
title: "L-GP Workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

load_ebsmoothr <- function() {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("../../EBSmoothr", quiet = TRUE, export_all = FALSE)
    return(invisible(TRUE))
  }
  if (requireNamespace("EBSmoothr", quietly = TRUE)) {
    library(EBSmoothr)
    return(invisible(TRUE))
  }
  stop("Install EBSmoothr or pkgload before rendering this document.")
}

load_ebsmoothr()
```

# Overview

This vignette illustrates one-dimensional empirical-Bayes smoothing with the
L-GP prior. The main entry points are:

- `LGP_setup()`, which builds the design and penalty matrices.
- `ebnm_LGP_generator()`, which returns the model-fitting function.

This vignette focuses on the known-standard-error generator workflow. If `s`
is unknown and you want to learn one common noise SD instead, see
`eb_smoother()` in `00-overview.Rmd`.

# Identity-Link Example

The identity link is the simplest setting: the posterior summaries target the
latent mean directly.

```{r}
set.seed(123)

t <- seq(0, 10, length.out = 150)
truth <- sin(t) + 0.3 * cos(1.5 * t)
s <- sample(c(0.15, 0.25, 0.4), length(t), replace = TRUE)
x <- truth + rnorm(length(t), sd = s)

setup_identity <- LGP_setup(
  t = t,
  p = 2,
  num_knots = 35,
  betaprec = 0,
  link = "identity"
)

fit_identity <- ebnm_LGP_generator(
  setup_identity,
  link = "identity"
)(x, s)

head(fit_identity$posterior)
fit_identity$fitted_g
```

```{r}
plot(t, x, pch = 16, cex = 0.45, col = "grey50",
     xlab = "t", ylab = "x",
     main = "L-GP with identity link")
lines(t, truth, col = "firebrick", lwd = 2, lty = 2)
lines(t, fit_identity$posterior$mean, col = "navy", lwd = 2)

upper <- fit_identity$posterior$mean + 2 * sqrt(fit_identity$posterior$var)
lower <- fit_identity$posterior$mean - 2 * sqrt(fit_identity$posterior$var)
polygon(c(t, rev(t)), c(upper, rev(lower)),
        border = NA, col = rgb(0, 0, 1, 0.15))
lines(t, fit_identity$posterior$mean, col = "navy", lwd = 2)
```

# Positive Mean Functions with the Log Link

When the mean function should remain positive, use the log link. In this case,
the posterior summaries are transformed back to the mean scale.

```{r}
set.seed(456)

truth_log <- exp(0.4 * sin(t) + 0.15 * cos(2 * t))
s_log <- rep(0.2, length(t))
x_log <- truth_log + rnorm(length(t), sd = s_log)

setup_log <- LGP_setup(
  t = t,
  p = 2,
  num_knots = 35,
  betaprec = 0,
  link = "log"
)

fit_log <- ebnm_LGP_generator(
  setup_log,
  link = "log"
)(x_log, s_log)
```

```{r}
plot(t, x_log, pch = 16, cex = 0.45, col = "grey50",
     xlab = "t", ylab = "x",
     main = "L-GP with log link")
lines(t, truth_log, col = "firebrick", lwd = 2, lty = 2)
lines(t, fit_log$posterior$mean, col = "darkgreen", lwd = 2)
```

# Role of `betaprec`

`betaprec` controls how the global polynomial coefficients are treated:

- `betaprec > 0`: proper Gaussian prior on the global coefficients.
- `betaprec = 0`: diffuse prior on the global coefficients.
- `betaprec < 0`: empirical-Bayes optimization mode for the global coefficients.

The following comparison shows how to create alternative setups:

```{r}
setup_diffuse <- LGP_setup(t = t, p = 2, num_knots = 30, betaprec = 0)
setup_proper  <- LGP_setup(t = t, p = 2, num_knots = 30, betaprec = 1)
setup_eb      <- LGP_setup(t = t, p = 2, num_knots = 30, betaprec = -1)

fit_diffuse <- ebnm_LGP_generator(setup_diffuse)(x, s)
fit_proper  <- ebnm_LGP_generator(setup_proper)(x, s)
fit_eb      <- ebnm_LGP_generator(setup_eb)(x, s)

c(
  diffuse = as.numeric(fit_diffuse$log_likelihood),
  proper = as.numeric(fit_proper$log_likelihood),
  eb = as.numeric(fit_eb$log_likelihood)
)
```

# Fixed Hyperparameters and Fixed Beta

When `fix_g = TRUE`, the L-GP fitter treats both the smoothing parameter and
the global coefficients as fixed. If `beta_fixed` is omitted, it defaults to a
zero vector.

```{r}
fit_fixed <- ebnm_LGP_generator(setup_eb)(
  x,
  s,
  g_init = LGP(0),
  fix_g = TRUE,
  beta_fixed = c(0.25, 0)
)

fit_fixed$fitted_g
fit_fixed$fitted_beta
```

# Useful Outputs

The L-GP fitter exposes several diagnostics that are helpful during method
development:

```{r}
names(fit_identity)
fit_identity$log_likelihood
fit_identity$log_likelihood_stepA
fit_identity$log_likelihood_stepB_joint
fit_identity$log_likelihood_stepB_laplace
```

In routine use, the most important components are usually:

- `fit$posterior$mean`
- `fit$posterior$var`
- `fit$fitted_g`
- `fit$log_likelihood`
