---
title: "Matern Workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)

# R
load_ebsmoothr <- function(path = NULL) {
  current_input <- if (requireNamespace("knitr", quietly = TRUE)) {
    knitr::current_input(dir = TRUE)
  } else {
    NULL
  }
  input_dir <- if (is.null(current_input)) NULL else dirname(current_input)
  candidate_paths <- unique(c(
    path,
    ".",
    "EBSmoothr",
    "../EBSmoothr",
    "../../EBSmoothr",
    if (is.null(input_dir)) NULL else file.path(input_dir, "../../EBSmoothr")
  ))

  is_ebsmoothr_source <- function(candidate) {
    desc <- file.path(candidate, "DESCRIPTION")
    if (!dir.exists(candidate) || !file.exists(desc)) {
      return(FALSE)
    }
    pkg_name <- tryCatch(unname(read.dcf(desc, fields = "Package")[1, 1]), error = function(e) NA_character_)
    isTRUE(pkg_name == "EBSmoothr")
  }

  source_loaded <- FALSE
  if (requireNamespace("pkgload", quietly = TRUE)) {
    for (candidate in candidate_paths) {
      if (is_ebsmoothr_source(candidate)) {
        pkgload::load_all(candidate, quiet = TRUE, export_all = FALSE)
        source_loaded <- TRUE
        break
      }
    }
  }

  if (!source_loaded) {
    if (!requireNamespace("EBSmoothr", quietly = TRUE)) {
      stop("Install EBSmoothr or pkgload before rendering this document.")
    }
    library(EBSmoothr)
  }
  if (!"backend" %in% names(formals(EBSmoothr::ebnm_Matern_generator))) {
    stop(
      "Loaded EBSmoothr is an older installed build without current Matern backend support. ",
      "Run this vignette from the EBSmooth project or reinstall EBSmoothr from the local source."
    )
  }
  invisible(TRUE)
}

load_ebsmoothr()
```

# Overview

`ebnm_Matern_generator()` builds an `ebnm`-compatible smoother using a
stationary Matern field represented through an INLA/SPDE mesh. The input
locations may be one-dimensional or two-dimensional.

This vignette focuses on the known-standard-error generator workflow. If `s`
is unknown and you want to learn one common noise SD instead, use
`eb_smoother()`. The high-level API supports `family = "matern", link = "log",
s = NULL`, so it can compare a non-negative smooth Matern prior against
`family = "constant", link = "log"`, which is the corresponding
non-negative constant-baseline prior.

The identity link uses the exact Gaussian backend by default. The log link is
available for positive mean functions and uses the Fisher Laplace backend by
default under `backend = "auto"`. Explicit `backend = "laplace"` keeps the
older observed-Hessian Laplace semantics.

There are multiple Matern backends:

- For `link = "identity"`, `backend = "auto"` uses the exact Gaussian EB
  backend. If `pc.penalty` is supplied, the PC prior contributes to the
  hyperparameter objective.
- If `link = "log"`, `backend = "auto"` resolves to `backend =
  "laplace_fisher"`. Fisher fits keep the same conditional mode as the
  log-link posterior objective, but use Fisher/Gauss-Newton curvature for the
  returned posterior approximation and score.
- Explicit `backend = "laplace"` uses observed-Hessian Laplace semantics.
  `backend = "laplace_tmb"` remains a compatibility alias for this route.
- Explicit `backend = "inla"` fits use INLA. The compatibility alias
  `backend = "inla_pc"` is accepted when `pc.penalty` is supplied, but the
  returned fit reports the canonical backend label `backend = "inla"`. For
  log-link fits, the primary comparable `log_likelihood` is the package
  observed-Hessian Laplace objective evaluated at the INLA mode.

# One-Dimensional Example

```{r}
set.seed(42)

locations_1d <- seq(0, 12, length.out = 120)
truth_1d <- exp(sin(locations_1d / 1.5) + 0.2 * cos(locations_1d))
s_1d <- sample(c(0.15, 0.25, 0.35), length(locations_1d), replace = TRUE)
x_1d <- truth_1d + rnorm(length(locations_1d), sd = s_1d)

fit_fun_1d <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "log"
)

fit_1d <- fit_fun_1d(x_1d, s_1d)

head(fit_1d$posterior)
fit_1d$fitted_g
fit_1d$fitted_beta
fit_1d$backend
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Matern smoother in one dimension")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 2)
lines(locations_1d, fit_1d$posterior$mean, col = "navy", lwd = 2)
```

# PC-Prior Fits

Supplying `pc.penalty` lets the Matern fitter select smoothing hyperparameters
under a PC prior. With `link = "identity"` and `backend = "auto"`, this remains
on the package exact Gaussian backend. Use explicit `backend = "inla"` when an
independent INLA/SPDE comparison is needed.

```{r}
fit_fun_pc <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "identity",
  pc.penalty = list(
    range = c(1.2, 0.5),
    sigma = c(0.6, 0.5)
  )
)

fit_pc <- fit_fun_pc(x_1d, s_1d)
pc_breakdown <- matern_objective_breakdown(fit_pc)

fit_pc$backend
fit_pc$fitted_g
fit_pc$fitted_beta
fit_pc$log_likelihood
fit_pc$log_likelihood_stepA_mlik_integration
pc_breakdown$objective_integrated_flat_plus_prior
pc_breakdown$recorded_minus_matched_exact
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Fisher Laplace and exact PC-prior Matern fits")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 1)
lines(locations_1d, fit_1d$posterior$mean, col = "navy", lwd = 2, lty = 2)
lines(locations_1d, fit_pc$posterior$mean, col = "darkorange3", lwd = 2, lty = 3)
legend(
  "topright",
  legend = c("truth", "Fisher Laplace", "exact PC prior"),
  col = c("firebrick", "navy", "darkorange3"),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

# Fixed Hyperparameters and Fixed Intercept

When `fix_g = TRUE`, the Matern smoother fixes the smoothing hyperparameters
from `g_init`. Supply `beta_fixed` separately when the intercept should also be
fixed.

```{r}
set.seed(99)

fit_fixed <- fit_fun_1d(
  x_1d,
  s_1d,
  g_init = Matern(theta = log(1.2), sigma = 0.7),
  fix_g = TRUE,
  beta_fixed = 0.1
)

fit_fixed$fitted_g
fit_fixed$fitted_beta
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Matern smoother with fixed hyperparameters")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 2)
lines(locations_1d, fit_fixed$posterior$mean, col = "darkgreen", lwd = 2)
```

Range/sigma fixing is not supported by the INLA backend. If `pc.penalty` is
supplied and you need fixed Matern hyperparameters, fit an unpenalized package
backend for that fixed-parameter score instead of requesting an INLA PC-prior
fit.

# Two-Dimensional Example

For a two-dimensional domain, pass an `n x 2` matrix of coordinates.

```{r}
set.seed(2026)

nx <- 14
ny <- 10
grid <- expand.grid(
  x = seq(0, 1, length.out = nx),
  y = seq(0, 1, length.out = ny)
)
locations_2d <- as.matrix(grid)

truth_2d <- with(
  grid,
  sin(2 * pi * x) * cos(2 * pi * y)
)
s_2d <- rep(0.15, nrow(locations_2d))
x_2d <- truth_2d + rnorm(nrow(locations_2d), sd = s_2d)

fit_fun_2d <- ebnm_Matern_generator(
  locations = locations_2d,
  max.edge = c(0.15, 0.3),
  link = "identity"
)

fit_2d <- fit_fun_2d(x_2d, s_2d)
```

```{r}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 5))

z_truth <- matrix(truth_2d, nrow = nx, ncol = ny)
image(
  x = sort(unique(grid$x)),
  y = sort(unique(grid$y)),
  z = z_truth,
  main = "True surface",
  xlab = "x",
  ylab = "y",
  col = hcl.colors(20, "BluYl")
)

z_fit <- matrix(fit_2d$posterior$mean, nrow = nx, ncol = ny)
image(
  x = sort(unique(grid$x)),
  y = sort(unique(grid$y)),
  z = z_fit,
  main = "Posterior mean",
  xlab = "x",
  ylab = "y",
  col = hcl.colors(20, "BluYl")
)

par(op)
```

# Useful Outputs

The Matern fitter returns backend-specific components in addition to the common
`ebnm`-style fields:

```{r}
names(fit_1d)
names(fit_pc)
class(fit_1d$mesh)
class(fit_1d$inla_result)
class(fit_pc$inla_result)
head(fit_1d$posterior_spatial_field)
```

The mesh is still useful for diagnostics and visualization. The
`posterior_spatial_field` component contains a latent-field summary for the
mesh nodes.

- In exact mode, `inla_result` is `NULL`, `posterior_sampler` is available, and
  the main `log_likelihood` is the exact EB marginal likelihood.
- In `pc.penalty` mode, `inla_result` stores the INLA fit object, the main
  `log_likelihood` is the package exact objective for identity-link fits or the
  package Laplace objective for log-link fits, both evaluated at the INLA mode.
  `posterior_sampler` is currently a compatibility stub that returns `NA`.
