---
title: "Matern Workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)

load_ebsmoothr <- function() {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("../../EBSmoothr", quiet = TRUE, export_all = FALSE)
    return(invisible(TRUE))
  }
  if (requireNamespace("EBSmoothr", quietly = TRUE)) {
    library(EBSmoothr)
    return(invisible(TRUE))
  }
  stop("Install EBSmoothr or pkgload before rendering this document.")
}

load_ebsmoothr()
```

# Overview

`ebnm_Matern_generator()` builds an `ebnm`-compatible smoother using a
stationary Matern field represented through an INLA/SPDE mesh. The input
locations may be one-dimensional or two-dimensional.

The current implementation is identity-link only.

There are now two Matern backends:

- If `pc.penalty = NULL`, the fitter uses the exact Gaussian EB backend and
  optimizes the exact marginal likelihood over `(log(range), log(sigma), beta0)`.
- If `pc.penalty` is supplied, the fitter switches to an INLA backend with
  `inla.spde2.pcmatern()`. In that case the main `log_likelihood` is the Step A
  penalized objective when `fix_g = FALSE`.

# One-Dimensional Example

```{r}
set.seed(42)

locations_1d <- seq(0, 12, length.out = 120)
truth_1d <- sin(locations_1d / 1.5) + 0.2 * cos(locations_1d)
s_1d <- sample(c(0.15, 0.25, 0.35), length(locations_1d), replace = TRUE)
x_1d <- truth_1d + rnorm(length(locations_1d), sd = s_1d)

fit_fun_1d <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "identity"
)

fit_1d <- fit_fun_1d(x_1d, s_1d)

head(fit_1d$posterior)
fit_1d$fitted_g
fit_1d$fitted_beta
fit_1d$backend
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Matern smoother in one dimension")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 2)
lines(locations_1d, fit_1d$posterior$mean, col = "navy", lwd = 2)
```

# PC-Prior Backend

Supplying `pc.penalty` switches the Matern fitter to an INLA backend. This is
useful when you want the smoothing hyperparameters to be selected under a PC
prior rather than under the exact EB objective.

```{r}
fit_fun_pc <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "identity",
  pc.penalty = list(
    range = c(1.2, 0.5),
    sigma = c(0.6, 0.5)
  )
)

fit_pc <- fit_fun_pc(x_1d, s_1d)
pc_breakdown <- matern_objective_breakdown(fit_pc)

fit_pc$backend
fit_pc$fitted_g
fit_pc$fitted_beta
fit_pc$log_likelihood
fit_pc$log_likelihood_stepA_mlik_integration
pc_breakdown$objective_integrated_flat_plus_prior
pc_breakdown$recorded_minus_matched_exact
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Exact EB and PC-prior Matern fits")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 2)
lines(locations_1d, fit_1d$posterior$mean, col = "navy", lwd = 2)
lines(locations_1d, fit_pc$posterior$mean, col = "darkorange3", lwd = 2)
legend(
  "topright",
  legend = c("truth", "exact EB", "INLA PC prior"),
  col = c("firebrick", "navy", "darkorange3"),
  lty = c(2, 1, 1),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

# Fixed Hyperparameters and Fixed Intercept

When `fix_g = TRUE`, the Matern smoother treats both the smoothing
hyperparameters and the intercept as fixed.

```{r}
set.seed(99)

fit_fixed <- fit_fun_1d(
  x_1d,
  s_1d,
  g_init = Matern(theta = log(1.2), sigma = 0.7),
  fix_g = TRUE,
  beta_fixed = 0.1
)

fit_fixed$fitted_g
fit_fixed$fitted_beta
```

```{r}
plot(locations_1d, x_1d, pch = 16, cex = 0.45, col = "grey55",
     xlab = "location", ylab = "x",
     main = "Matern smoother with fixed hyperparameters")
lines(locations_1d, truth_1d, col = "firebrick", lwd = 2, lty = 2)
lines(locations_1d, fit_fixed$posterior$mean, col = "darkgreen", lwd = 2)
```

If you supply both `fix_g = TRUE` and `pc.penalty`, the returned
`log_likelihood` is a penalized pseudo-objective:

```text
Step B conditional marginal likelihood + log PC prior on (log(range), log(sigma)).
```

```{r}
fit_pc_fixed <- fit_fun_pc(
  x_1d,
  s_1d,
  g_init = Matern(theta = log(1.2), sigma = 0.7),
  fix_g = TRUE,
  beta_fixed = 0.1
)

fit_pc_fixed$log_likelihood
fit_pc_fixed$log_likelihood_stepB
fit_pc_fixed$log_likelihood_pc_prior_theta
```

# Two-Dimensional Example

For a two-dimensional domain, pass an `n x 2` matrix of coordinates.

```{r}
set.seed(2026)

nx <- 14
ny <- 10
grid <- expand.grid(
  x = seq(0, 1, length.out = nx),
  y = seq(0, 1, length.out = ny)
)
locations_2d <- as.matrix(grid)

truth_2d <- with(
  grid,
  sin(2 * pi * x) * cos(2 * pi * y)
)
s_2d <- rep(0.15, nrow(locations_2d))
x_2d <- truth_2d + rnorm(nrow(locations_2d), sd = s_2d)

fit_fun_2d <- ebnm_Matern_generator(
  locations = locations_2d,
  max.edge = c(0.15, 0.3),
  link = "identity"
)

fit_2d <- fit_fun_2d(x_2d, s_2d)
```

```{r}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 5))

z_truth <- matrix(truth_2d, nrow = nx, ncol = ny)
image(
  x = sort(unique(grid$x)),
  y = sort(unique(grid$y)),
  z = z_truth,
  main = "True surface",
  xlab = "x",
  ylab = "y",
  col = hcl.colors(20, "BluYl")
)

z_fit <- matrix(fit_2d$posterior$mean, nrow = nx, ncol = ny)
image(
  x = sort(unique(grid$x)),
  y = sort(unique(grid$y)),
  z = z_fit,
  main = "Posterior mean",
  xlab = "x",
  ylab = "y",
  col = hcl.colors(20, "BluYl")
)

par(op)
```

# Useful Outputs

The Matern fitter returns backend-specific components in addition to the common
`ebnm`-style fields:

```{r}
names(fit_1d)
names(fit_pc)
class(fit_1d$mesh)
class(fit_1d$inla_result)
class(fit_pc$inla_result)
head(fit_1d$posterior_spatial_field)
```

The mesh is still useful for diagnostics and visualization. The
`posterior_spatial_field` component contains a latent-field summary for the
mesh nodes.

- In exact mode, `inla_result` is `NULL`, `posterior_sampler` is available, and
  the main `log_likelihood` is the exact EB marginal likelihood.
- In `pc.penalty` mode, `inla_result` stores the INLA fit object, the main
  `log_likelihood` is backend-dependent, and `posterior_sampler` is currently a
  compatibility stub that returns `NA`.
