---
title: "Package Sanity Checks"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

load_ebsmoothr <- function() {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("../../EBSmoothr", quiet = TRUE, export_all = FALSE)
    return(invisible(TRUE))
  }
  if (requireNamespace("EBSmoothr", quietly = TRUE)) {
    library(EBSmoothr)
    return(invisible(TRUE))
  }
  stop("Install EBSmoothr or pkgload before rendering this document.")
}

load_ebsmoothr()

check_results <- list()

record_check <- function(name, expr) {
  ok <- isTRUE(expr)
  if (!ok) {
    stop(sprintf("Sanity check failed: %s", name))
  }
  check_results[[length(check_results) + 1L]] <<- data.frame(
    check = name,
    status = "PASS",
    stringsAsFactors = FALSE
  )
  invisible(TRUE)
}

collect_checks <- function() {
  do.call(rbind, check_results)
}
```

# Purpose

This internal vignette is a light end-to-end sanity check for the current
`EBSmoothr` package interface. It is not a replacement for the `testthat`
suite. The goal is simpler:

- verify that the main public fitting paths still run;
- verify that the returned objects have the expected structure;
- verify a few key invariants for the newer exact and PC-prior Matern backends.

# L-GP Identity Link

```{r}
set.seed(1001)

t <- seq(0, 8, length.out = 80)
truth_lgp <- sin(t) + 0.25 * cos(1.5 * t)
s_lgp <- rep(0.2, length(t))
x_lgp <- truth_lgp + rnorm(length(t), sd = s_lgp)

setup_lgp_identity <- LGP_setup(
  t = t,
  p = 2,
  num_knots = 25,
  betaprec = 0,
  link = "identity"
)

fit_lgp_identity <- ebnm_LGP_generator(
  setup_lgp_identity,
  link = "identity"
)(x_lgp, s_lgp)

record_check(
  "L-GP identity returns one posterior row per observation",
  nrow(fit_lgp_identity$posterior) == length(t)
)
record_check(
  "L-GP identity posterior means are finite",
  all(is.finite(fit_lgp_identity$posterior$mean))
)
record_check(
  "L-GP identity log_likelihood is finite",
  is.finite(as.numeric(fit_lgp_identity$log_likelihood))
)

fit_lgp_identity$fitted_g
head(fit_lgp_identity$posterior)
```

# L-GP Log Link and Fixed Beta

```{r}
set.seed(1002)

truth_lgp_log <- exp(0.35 * sin(t) + 0.1 * cos(2 * t))
s_lgp_log <- rep(0.18, length(t))
x_lgp_log <- truth_lgp_log + rnorm(length(t), sd = s_lgp_log)

setup_lgp_log <- LGP_setup(
  t = t,
  p = 2,
  num_knots = 25,
  betaprec = -1,
  link = "log"
)

fit_lgp_log <- ebnm_LGP_generator(
  setup_lgp_log,
  link = "log"
)(x_lgp_log, s_lgp_log)

fit_lgp_fixed <- ebnm_LGP_generator(setup_lgp_log)(
  x_lgp,
  s_lgp,
  g_init = LGP(0),
  fix_g = TRUE,
  beta_fixed = c(0.2, 0)
)

record_check(
  "L-GP log-link posterior means are positive",
  all(fit_lgp_log$posterior$mean > 0)
)
record_check(
  "L-GP fixed-beta path returns the requested beta",
  max(abs(fit_lgp_fixed$fitted_beta - c(0.2, 0))) < 1e-8
)

c(
  lgp_log_likelihood = as.numeric(fit_lgp_log$log_likelihood),
  lgp_fixed_likelihood = as.numeric(fit_lgp_fixed$log_likelihood)
)
```

# Exact Matern in One Dimension

```{r}
set.seed(1003)

locations_1d <- seq(0, 10, length.out = 90)
truth_matern_1d <- 0.3 + sin(locations_1d / 1.8)
s_matern_1d <- rep(0.22, length(locations_1d))
x_matern_1d <- truth_matern_1d + rnorm(length(locations_1d), sd = s_matern_1d)

fit_matern_exact_1d <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "identity"
)(x_matern_1d, s_matern_1d)

record_check(
  "Exact Matern backend label is correct",
  identical(fit_matern_exact_1d$backend, "exact")
)
record_check(
  "Exact Matern inla_result is NULL",
  is.null(fit_matern_exact_1d$inla_result)
)
record_check(
  "Exact Matern posterior means are finite",
  all(is.finite(fit_matern_exact_1d$posterior$mean))
)
record_check(
  "Exact Matern fitted sigma is positive",
  fit_matern_exact_1d$fitted_g$sigma > 0
)

fit_matern_exact_1d$fitted_g
fit_matern_exact_1d$fitted_beta
```

# Exact Matern in Two Dimensions

```{r}
set.seed(1004)

grid_2d <- expand.grid(
  x = seq(0, 1, length.out = 7),
  y = seq(0, 1, length.out = 6)
)
locations_2d <- as.matrix(grid_2d)
truth_matern_2d <- with(grid_2d, sin(2 * pi * x) * cos(2 * pi * y))
s_matern_2d <- rep(0.2, nrow(locations_2d))
x_matern_2d <- truth_matern_2d + rnorm(nrow(locations_2d), sd = s_matern_2d)

fit_matern_exact_2d <- ebnm_Matern_generator(
  locations = locations_2d,
  max.edge = c(0.25, 0.4),
  link = "identity"
)(x_matern_2d, s_matern_2d)

record_check(
  "Exact Matern 2D posterior length matches the data length",
  nrow(fit_matern_exact_2d$posterior) == nrow(locations_2d)
)
record_check(
  "Exact Matern 2D latent-field summary is non-empty",
  nrow(fit_matern_exact_2d$posterior_spatial_field) > 0
)

cor(
  fit_matern_exact_2d$posterior$mean,
  truth_matern_2d
)
```

# Matern with `pc.penalty`

```{r}
set.seed(1005)

fit_fun_matern_pc <- ebnm_Matern_generator(
  locations = locations_1d,
  link = "identity",
  pc.penalty = list(
    range = c(1.0, 0.5),
    sigma = c(0.7, 0.5)
  )
)

fit_matern_pc <- fit_fun_matern_pc(x_matern_1d, s_matern_1d)
fit_matern_pc_fixed <- fit_fun_matern_pc(
  x_matern_1d,
  s_matern_1d,
  g_init = Matern(theta = log(1.0), sigma = 0.7),
  fix_g = TRUE,
  beta_fixed = 0.3
)

record_check(
  "PC-prior Matern auto backend remains exact",
  identical(fit_matern_pc$backend, "exact")
)
record_check(
  "PC-prior Matern auto fit does not store an INLA result",
  is.null(fit_matern_pc$inla_result)
)
record_check(
  "PC-prior Matern stores a finite PC-prior contribution",
  is.finite(fit_matern_pc$log_likelihood_pc_prior_theta)
)
record_check(
  "PC-prior fixed mode stores a finite objective",
  is.finite(as.numeric(fit_matern_pc_fixed$log_likelihood))
)

c(
  pc_log_likelihood = as.numeric(fit_matern_pc$log_likelihood),
  pc_prior_contribution = fit_matern_pc$log_likelihood_pc_prior_theta,
  pc_fixed_penalized = fit_matern_pc_fixed$log_likelihood
)
```

# Summary Table

```{r}
knitr::kable(
  collect_checks(),
  align = c("l", "c"),
  caption = "Summary of package-level sanity checks."
)
```

# Takeaway

This vignette confirms that the current package can still run its main public
paths:

- L-GP with identity and log links;
- L-GP with fixed smoothing parameter and fixed beta coefficients;
- exact Matern in one and two dimensions;
- Matern with `pc.penalty`, both in Step A optimization mode and in fixed
  hyperparameter mode.

For stricter regression coverage, the authoritative checks remain the
`testthat` suite in `EBSmoothr/tests/testthat`.
